### 2010-08-27 Thesis/Chapter6.tex
- Update 27 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@167 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
A short abstract for the upcoming chapter
\end{shortAbstract}

@@

Eventually, vertex and pixel shaders became programmable to enable game programmers to generate even more realistic effects. Programmable pixel shaders allow the programmer to substitute, for example, a lighting model other than those provided by default by the graphics card. Shaders have enabled graphics programmers to create lens effects, displacement mapping, and depth of field. This evolution of GPU's hardware and the increasing programmable capability naturally lead to use GPUs for non-graphics applications. The term GPGPU (General-purpose computation on graphics processing units) was coined by Mark Harris in 2002 when he recognised an early trend of using GPUs for non-graphics applications. However, capabilities of GPU's were still fairly limited for non-graphics applications.

@@

GPUs may be seen as high-performance many-core processors that can be used to accelerate a wide range of applications. An example (given by Sanford Russell from NVIDIA) to think about to illustrate the difference between a traditional CPU and a GPU is this: if you were looking for a word in a book, and handed the task to a CPU, it would start at page $ 1 $ and read it all the way to the end, because it's a serial processor. It would be fast, but would take time because it has to go in order. A GPU, which is a parallel processor, would tear the book into a thousand pieces and read it all at the same time. Even if each individual word is read more slowly, the book may be read in its entirety quicker, because words are read simultaneously.

In addition to an increasing programmability, it is easy to understand why the development of GPGPU is soaring. GPU is now used in imaging, finance, signal processing, simulation, video and many other fields. A recent review of GPGPU algorithms may be found in \cite{Owens07}. This is therefore not surprising that research has been carried out to leverage the power of GPUs to solve partial differential equations of continuum mechanics via the computationally expensive finite element method. Indeed, we have already insisted on the strong time constraints demanded by the field of medical simulation and the path towards GPGPU is natural. A non-exhaustive review of algorithms implemented on GPU in the context of medical simulation was given chapter \ref{chap4} (page~\pageref{chap4:GPUMedicalSimulation}).

@@

		\subsubsection*{Graphics API}

		\subsubsection*{Non-graphics API}

\section{Implementation into SOFA}
	\subsection{Presentation of SOFA project and architecture}
	\subsection{CUDA description (why CUDA over Cg?)}	
	\subsection{GPU implementation of TLED in SOFA and release in open-source \OC{ISBMS 2008a}}

\section{Results}	
	\subsection{Pure shear of a cube}
	\subsection{Compression of a cube}
	\subsection{GPU performance}
	\subsection{Simulation of liver deformation}
	\subsection{A medical application: cataract surgery}

>>>>

\chapter{GPU implementation of TLED}
\label{chap6}
\begin{shortAbstract}
\OC{ISBMS 2008a}
\end{shortAbstract}

@@

Eventually, vertex and pixel shaders became programmable to enable game programmers to generate even more realistic effects. Programmable pixel shaders allow the programmer to substitute, for example, a lighting model other than those provided by default by the graphics card. Shaders have enabled graphics programmers to create lens effects, displacement mapping and depth of field. This evolution of GPU's hardware and the increasing programmable capability naturally lead to use GPUs for non-graphics applications. The term GPGPU (General-purpose computation on graphics processing units) was coined by Mark Harris in 2002 when he recognised an early trend of using GPUs for non-graphics applications. However, capabilities of GPU's were still fairly limited for non-graphics applications at this time.

@@

GPUs may be seen as high-performance many-core processors that can be used to accelerate a wide range of applications. An example (given by Sanford Russell from NVIDIA) to think about to illustrate the difference between a traditional CPU and a GPU is this: if you were looking for a word in a book, and handed the task to a CPU, it would start at page 1 and read it all the way to the end, because it's a serial processor. It would be fast, but would take time because it has to go in order. A GPU, which is a parallel processor, would tear the book into a thousand pieces and read it all at the same time. Even if each individual word is read more slowly, the book may be read in its entirety quicker, because words are read simultaneously.

In addition to an increasing programmability, it is easy to understand why the development of GPGPU is soaring. GPU is now used in imaging, finance, signal processing, simulation, audio and video processing, astronomy, weather forecasting, molecular modelling, cryptography, quantum mechanics and many other fields. A fairly recent review of GPGPU algorithms may be found in \cite{Owens07}. This is therefore not surprising that research has been carried out to leverage the power of GPUs to solve partial differential equations of continuum mechanics via the computationally expensive finite element method. Indeed, we have already insisted on the strong time constraints demanded by the field of medical simulation and the path towards GPGPU is natural. A non-exhaustive review of algorithms implemented on GPU in the context of medical simulation was given chapter \ref{chap4} (page~\pageref{chap4:GPUMedicalSimulation}).

@@

Interface with a GPU has traditionally been via a graphic application programming interface (API) such as \citetalias{Board05} or \citetalias{DirectX10}. OpenGL is an open standard API that provides a number of functions for the rendering of 2D and 3D graphics and is available on most modern operating systems including but not limited to Windows, Mac OS X and Linux. It was initially designed by Silicon Graphics Inc. and now managed by the technology consortium Khronos group. DirectX is a proprietary API that provides functions to render three dimensional graphics, and uses hardware acceleration if it is available on the graphics card. It was designed by Microsoft for use on the Windows platform.

Because OpenGL and DirectX are both low-level APIs and include many irrelevant functionality from a GPGPU perspective, higher level languages were designed to give developers more direct control of the graphics pipeline without having to use assembly language or hardware-specific languages.

		\subsubsection*{Graphics API}
OpenGL (version 1.5 and newer) provides shader language based on the C programming language called OpenGL shading language (GLSL). In the DirectX API (DirectX 9 and newer), shaders are programmed with the high level shader language (HLSL). Cg is another high-level shading language developed by NVIDIA in close collaboration with Microsoft and it is very similar to Microsoft's HLSL. However, the Cg compiler is not specific to a graphics API and can output both OpenGL and DirectX shader programs.

These languages all share basic data types (float, integer, bool etc.) and also feature vector and matrix data types that are based on the basic data types (like float3 or float4x4 in Cg and vec3 or mat4 in GLSL) and standard library functions are provided (vector and matrix multiplication, cross product etc.). In addition, they support loops and branching like if/else, while, and for. However, because these languages are just an abstraction on the top of graphics API, there are significant restrictions on what they can do. Specifically, shader programs are unable to determine the destination of their outputs. They merely operate on fragment data (color, texture coordinates, normal etc.) and return the modified fragment at the same location. Each shader is also executed independently of its neighbours since no communication is allowed between processors. Note that while this was made possible on recent hardware, these graphics API do not take advantage of this feature and therefore this statement remains valid.

Memory storage is allowed through the use of textures, which essentially are structured collections of floating point values arranged as arrays. Shaders can read textyres from any location and as many time as they like. As such it is a concept very similar to regular arrays. However, the individual components of textures consist of up to four floating point values in the form of a vector. This comes from the fact that graphics processing units were initially designed to manipulate color images, so each component of a texture is in fact a vector holding the red, green, blue and alpha (transparency) values for the given pixel. Textures are therefore optimised for retrieving up to four floating point values in a single read.

When it comes to writing to GPU memory though, things are a bit more limited. Indeed, in contrast to texture reads, shaders can only write once into textures (at the end of the set of instructions). Morevoer, as stated earlier, shaders can only write in predefined locations corresponding to their potential screen location. Therefore, graphics API does not allow writing to random memory locations (this feature is called \emph{scattering}). Nethertheless, shaders can have multiple render targets. The write locations within each texture are still fixed though.

		\subsubsection*{Non-graphics API}
Eventually, non-graphics API were released to support the development of GPGPU. The two major graphics card manufacturers first released their own API. While the objective of hiding graphics aspects is the same, the approaches of NVIDIA and ATI (now owned by AMD) are quite different.

In November 2006, NVIDIA released CUDA which consists of C language extensions. CUDA is fairly high level. Programs (called kernels) may be coded as C functions and invoked directly without going through a render pass as with graphics APIs. When called, kernels are executed N times in parallel by N threads as opposed to only once like regular C functions. Memory management functions analogous to standard C are also provided (memcpy, free etc.).

At the same time, ATI released CTM released a software development layer aimed for low-level access (assembly-style language). Memory management functions similar to graphics API are also provided. A year later, in December 2007, ATI released Stream SDK and added a new high-level language derived from C, called ATI Brook+ which later evolved to ATI CAL (Compute Abstraction Layer). In both cases, these APIs work solely on each manufacturer's devices, CUDA with NVIDIA cards and Stream SDK on ATI cards.

OpenCL is a more general framework for writing programs that execute across heterogeneous platforms consisting of CPUs, GPUs, and other processors. It is not specific to a single type of hardware. OpenCL was initially developed by Apple and refined into an initial proposal in collaboration AMD (ATI), IBM, Intel, and NVIDIA. Apple submitted this initial proposal to the Khronos Group and the first release occurred in December 2008. AMD decided to support OpenCL instead of the now deprecated CTM in its Stream SDK.

The latest API to be released was DirectCompute from Microsoft. DirectCompute was released as part of DirectX11 in October 2009 and allows access to general-purpose computing with a fairly low level language.

	\subsubsection*{The choice of API}	
The first GPU implementation	of the TLED algorithm was carried out by \cite{Taylor07b,Taylor08}. They used the Cg language and had to reformulate the TLED algorithm presented to accomodate with the limited possibilities of this graphics API. Indeed, the major difference between this parallel implementation and a serial one is the writing of element nodal forces to memory at the end of a first kernel and their subsequent retrieval and summation during a second kernel. In a serial implementation nodal force contributions would most likely be added to a global sum as they are computed, rather than stored and summed in a second loop. But as stated earlier, this graphics API does not allow scattering, that is writing to random memory locations. Because of this limitation, element nodal forces cannot be directly added to the total nodal forces. Instead, the authors had to reformulate force summation as a gather operation in their implementation. They reported significant solution speed gains up to $ 16.8 \times$ over the CPU implementation. Thus, they were able to compute a simple cube model with up to $ 16\,000 $ tetrahedral elements in real-time.

A major improvement was introduced with the release of CUDA. In theory, CUDA allows scattered writes and this new feature could dramatically change the GPU implementation of the TLED as the reformulation of the scatter as a gather would no longer be useful. A single kernel would be sufficient and an increase in performance is expected. Along with a more readable and simple code, this is the main feature that made us choose the CUDA API to reimplement the TLED algorithm.

In addition, we decided to develop our CUDA-based re-implementation of the TLED algorithm within the international and open source framework SOFA. We will show that this integration has a very limited cost in terms of performance by comparing the SOFA version with a standalone implementation. By providing an efficient and accurate non-linear FEM for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

\section{Implementation into SOFA}

	\subsection{SOFA, an open source simulation framework}

\subsubsection*{Objectives}
The multi-disciplinary aspect of medical simulation requires the integration within a single environment of solutions in areas as diverse as visualisation, biomechanical modelling, haptic feedback and contact modelling. Although their interaction is essential to design a realistic simulator, only few teams have the sufficient resources to build such frameworks. This diversity of problems creates challenges for researchers to advance specific areas, and leads rather often to duplication of effort. The open source SOFA framework \citep{Allard07} was created to overcome this issue by providing researchers with an advanced software architecture that facilitates the development of new algorithms and simulators. SOFA has been mostly developed by INRIA (the French national institute for research in computer science and control) and CIMIT (Center for Integration of Medicine and Innovative Technology) and is primarily targeted at real-time simulation with an emphasis on medical simulation. SOFA is highly modular and flexible: it allows independently developed algorithms to interact together within a common simulation while minimising the development time required for integration. The overall goal is to develop a flexible framework while minimising the impact of this flexibility on the computation overhead. To achieve these objectives, SOFA proposes a new architecture that implements a series of concepts described below.

\subsubsection*{SOFA architecture}

\paragraph{High-level modularity.}
The SOFA architecture relies on the innovative notion of multi-model representation where an object is explicitly decomposed into various representations: Behaviour Model, Collision Model, Collision Response Model, Visual Model and Haptic Model. Each representation can then be optimised for a particular task (biomechanics,  collision detection, visualisation, haptics) while at the same time improving interoperability by creating a clear separation between the functional aspects of the simulation components. These representations are then connected together via a mechanism called \emph{mapping}. Various mapping functions can be defined, and each mapping associates a set of primitives of a representation to a set of primitives in the other representation (\fig{chap6:fig-SOFA}). For instance, a mapping can connect degrees of freedom in a Behaviour Model to vertices in a Visual Model.

\begin{figure}
\includegraphics[width=13.9cm]{chapter6/sofa.pdf}
\caption [Multi-model representation in SOFA] {Multi-model representation in SOFA. \textbf{Left:} a Behaviour Model controls the other representations via a series of mappings. \textbf{Right:} examples of representations with a liver model.}
\label{chap6:fig-SOFA}
\end{figure}

\paragraph{Finer level modularity.}
In order to easily compare algorithms within SOFA, more flexibility was added to the Behaviour Model by introducing an even finer level of granularity. A series of generic primitives common to most physics-based simulations have been defined: DoF, Mass, Force Field and Solver. The DoF component describes the degrees of freedom, and their derivatives, of the object. The Mass component represents its mass. The Force Field describes both internal and external forces that can be applied to this object. The Solver component handles the time step integration, i.e. advancing the state of the system from time $t$ to time $t+\Delta t$.

\paragraph{Scene-graph.}
Finally, another key aspect of SOFA is the use of a scene-graph to organise and process the elements of a simulation. Each component is attached to a node of a tree structure. This simple structure makes it easy to visit all or a subset of the components in a scene, and dependencies between components are handled by retrieving sibling components attached to the same node. 
During the simulation loop, most computations can be expressed as a traversal of the scene-graph. For instance, at each time step, the simulation state is updated by processing all Solver components, which will then forward requests to the appropriate components by recursively sending actions within its sub-tree.

These different functionalities and levels of abstraction allow the user to switch from one component to another by simply editing an XML file, without having to recompile. In particular this permits testing of different computational models of soft tissue deformation, and to assess the pros and cons of various algorithms within the same context.

	\subsection{CUDA description}
GPUs achieve a high floating point capacity by distributing computation across a high number of parallel execution threads. They perform optimally as single instruction, multiple data devices. CUDA is a relatively new C API for compatible NVIDIA GPUs. CUDA organises threads in two hierarchical levels: blocks, which are groups of threads executed on one of the GPU's multiprocessors, and grids, which are groups of blocks launched concurrently on the device, and which all execute the same kernel. Figure~\ref{chap6:fig-cuda} represents this thread organisation. 
% As an example, the NVIDIA 8800 GTX used for the results presented in section \ref{results_section} has 16 multiprocessors, each containing eight processors.

\begin{figure}
\begin{center}
\includegraphics[width=10cm]{chapter6/cuda_architecture.pdf} 
\caption[CUDA architecture] {Each kernel is executed by CUDA as a group of threads within a grid. Image courtesy of NVIDIA. }
\label{chap6:fig-cuda}
\end{center}
\end{figure}

CUDA allows developers to specify the number of threads per block in each execution (the so-called execution configuration), effectively defining the distribution of computational load across all processors. For a given kernel the block dimensions are chosen to optimise the utilisation of the available computational resources. Care should be taken at the multiprocessor level in balancing the available memory required by the kernels with the ability to hide global memory latency. Since a finite amount of memory is available on a multiprocessor, the memory requirements of a kernel will determine how many threads can run concurrently on each. Importantly, CUDA's use of time slicing allows more than one block to be executed concurrently on a single multiprocessor, which has important implications for hiding memory latency. If more than one block is executing, the multiprocessor is able to switch processing between blocks while others are stalled on memory accesses, whereas it has no option but to wait for these if only one block is executing. Therefore for memory bandwidth bound kernels it may be preferable to launch several smaller blocks on each multiprocessor rather than a single larger one if both configurations make the same use of multiprocessor memory resources. While tools are available from NVIDIA for estimating the optimal execution configuration, it has proved necessary to fine tune the configuration experimentally for each kernel.

	\subsection{First implementation: scatter as a gather}

\subsubsection*{SOFA integration.}
Implementing a biomechanical model in SOFA translates essentially into writing a new Force Field, i.e. describing the algorithm used to compute internal forces in the model. It merely comes down to creating a single C++ class and changing the position reads and force writes to integrate the algorithm into SOFA's design. The precomputation phase takes place in the initialisation method where relevant variables are computed and passed to an external C function that allocates memory on the GPU and binds textures to it. During the simulation loop, the Solver requests the computation of the forces by launching the appropriate kernels on the GPU.

\subsubsection*{Kernel organisation.}
Although CUDA allows scattered writes in theory, it offers no write conflict management between threads. Why is this a problem? The major steps of the TLED algorithm are the following: (1) we compute the element nodal forces, that is for each node of a given element, we calculate the force contribution that this element has on this node; (2) we obtain the global force for each node by summing up the force contributions of all elements to which this node is attached; (3) we compute the displacement for each node by integration using the central difference method. In a serial implementation (CPU) we would do a loop on all elements and for each element we compute all element nodal force contributions. We would then directly add the element force contributions computed for each node into a global node array at the corresponding location for each node of the current processed element. In a parallel implementation, many elements are processed concurrently, in parallel. Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their computed nodal force contribution into the exact same location in GPU's memory. At the time of this first implementation (2007), the result of concurrent writes was undefined. In fact, only one write was guaranteed to happen and there was no way to know which one succeeded. Potential measures to address this have proved extremely detrimental to performance. For this reason, the kernel arrangement in our first CUDA implementation eventually ends up to be essentially the same as the one used by \cite{Taylor07b}.

Consequently, the TLED CUDA implementation also relies on 2 kernels. The first kernel operates over elements in the model and computes the element stresses based on the current model configuration. It then converts these into nodal force contributions, which are written to global memory. The second kernel operates over nodes and reads the previously calculated element force contributions and sums them for each node. The SOFA central difference solver computes the displacements from the nodal forces. Therefore, due to the impracticability of scattered writes, the sum operation is reformulated as a gather and the second kernel is needed to sum the nodal forces. The use of the developed viscoelastic constitutive update scheme necessitates storage of an additional array of state variables $ \boldsymbol \Upsilon_i $.

\subsubsection*{Memory usage.}
One efficient method for reading global memory data within kernels is texture fetching. Textures may be bound to either \emph{cudaArrays} or regions of linear memory. CudaArrays have been designed to achieve optimal fetching when the access pattern has a high level of 2D locality. In the present application, the access pattern among threads is essentially random (since unstructured meshes are used) and our experiments have shown that texture fetching from linear memory is in fact fastest. Therefore all global memory precomputed variables were accessed using this method.

In SOFA, the forces are stored on the GPU in global memory. Since this memory space is not cached, it is important to follow the appropriate access pattern to obtain maximum memory bandwidth, especially given how costly accesses to device memory are. A multiprocessor takes 4 clock cycles to issue one memory instruction for a set of threads. When accessing global memory, there are, in addition, 400 to 600 clock cycles of memory latency. A suboptimal access pattern would yield incoherent writes. The memory bandwidth would then be an order of magnitude lower. In order to prevent this, a key feature of CUDA has been used: \emph{shared memory}. This is a very fast memory shared by all the processors of a multiprocessor. Hence, the results of the second kernel are first copied to shared memory and then moved to global memory. If the copies are well organised, it is possible to re-order the access to fulfil all the memory requirements (for both shared and global) and thus reach the maximum bandwidth.

\subsubsection*{CPU-GPU interaction.}
CPU-GPU interaction is generally a significant bottleneck in General Purpose GPU applications due to the relatively low interface bandwidth and it is desirable to minimise such interaction. However, interaction cannot be entirely removed from the present implementation since, for example, the solver requires inputs in the form of loaded nodes (which may change due to the interaction with the user) and their displacements, and may need to provide outputs in the form of reaction forces for haptic feedback. CUDA alleviates the problem somewhat by allowing allocation of areas of page-locked host memory which are directly accessible by the GPU and therefore offer much higher bandwidth. In SOFA, all transfers between CPU and GPU are made via this mechanism.

\subsubsection*{Element technology.}
We used both reduced integration 8-node hexahedral and 4-node (linear) tetrahedral elements. Tetrahedral meshes are easily generated and therefore widely used in simulations. However, hexahedra are preferable both in terms of solution accuracy and computational efficiency. Indeed, 4-node linear tetrahedra are known to be susceptible to volumetric locking \citep{Hughes00}.  Yet, a disadvantage of hexahedra is the difficulty in automatically generating hexahedral meshes over arbitrary domains. Construction of hexahedral meshes is time consuming and laborious, which fact is of even greater significance when patient-specific simulations are considered. For this reason tetrahedral meshes are widely used in simulations despite their short comings.

Another drawback of using hexahedra is the existence of so-called Hourglass modes that have to be addressed to avoid deterioration of the solution \citep{Flanagan81}. Techniques for suppressing these modes exist, but naturally involve additional computations. However, for a given number of degrees of freedom (DOF), a hexahedral mesh can be built with far fewer elements than a tetrahedral one. Since the majority of the calculations in explicit dynamic analyses are performed per element, this results in reduced overall computation time.

From a GPU perspective, hexahedral element computations are substantially heavier and demand more memory resources than linear tetrahedral elements. Most of the matrices (such as shape function derivatives and nodal displacements) are twice as large for hexahedra, which necessitates twice as many texture fetches per element, and use of twice as many registers per thread. Similarly, twice as many nodal forces per element are written to global memory. Additional variables associated with hourglass control are required also. Therefore, on a per element basis hexahedra are significantly less efficient than tetrahedra, especially for GPU execution where memory efficiency is crucial. Thus, we observe that the occupancy (GPU percentage usage) drops from $25\,\%$ to only $8\,\%$ when using hexahedral elements. However, from the point of view of an entire model the lower number of hexahedra required for a given number of degrees of freedom still outweighs this element-wise inefficiency.

	\subsection{Second implementation: a better use of shared memory}

\subsubsection*{Limitations of the previous implementation}
The previous TLED algorithm has been extensively tested on regular meshes like cubes. The model was purposely simple to test our new CUDA implementation. However, the TLED algorithm was developed to compute the deformation of organs. Thus, we started to assess our TLED implementation on an irregular mesh: a liver. From a segmented liver obtained from IRCAD (Research institute against digestive system cancer, France), we meshed the surface using a marching cube technique \citep{Lorensen87}. After smoothing it with Blender, a tetrahedral mesh has been generated with Tetgen, a free mesh generator. From there, we realised that the TLED was highly non efficient with such a mesh. After an extensive analysis with the profiler provided by NVIDIA, we became able to explain it. The liver mesh has a higher valency, meaning that each node has more elements attached than within a regular mesh like a cube. And on GPU, nodes are processed in parallel by group of 32 threads (called a warp). But if the instructions are different within a warp, the warp is divergent and the computations are serialised. With its higher valency, the liver mesh involves much more divergent warps when the forces are added up. To give an idea of the difference, a cube ($343$ nodes and $1\,296$ elements) is computed in $0.16\,$ms when a liver mesh ($399$ nodes and $1\,250$ elements) is processed in $0.27\,$ms. Based on the limitations discovered when the TLED has been applied to this irregular mesh, we came up with a new design, that we implemented and tested.

\subsubsection*{A single kernel approach}
The main bottleneck of the previous implementation was adding up the forces on each node in a second kernel after the computation of element force contributions in a first kernel. The latter were stored in 4 different textures (8 for the hexahedral formulation) and in order to add up the forces the second kernel had to switch all the time between the correct textures within the warps, causing them to diverge and slowing down substantially the overall process. Our idea was therefore to use shared memory to avoid writing in global memory at the end of the first kernel and prevent the warps to diverge. As explained before, shared memory is a very fast memory shared by all the processors of a multiprocessor. And each block of threads is sent to a multiprocessor to be executed. Obviously the amount of shared memory is finite, which limits the validity of this new approach and the results in performance will depend on the hardware (which may have different amount of shared memory).

In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks). However, storing the force contributions in shared memory instead of global memory to read them later on via texture fetches widely outperforms the need of computing several elements several times.

\TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}

	\subsection{Third implementation: atomic writes}
In April 2010, NVIDIA released a family of graphics cards based on a new GPU architecture called Fermi. Among many new features and improvements over previous generation, these cards added the capability of so-called \emph{atomic writes} for floats. The write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads. In other words, no other thread can access this address until the operation is complete. The concurrent writes are simply serialised in hardware. This feature allows us to implement the TLED algorithm as we would on a CPU, using a single kernel for all computations, and without the need to re-order the nodes and elements organisation as in the second implementation.

\TODO{If time, try simple implementation using atomic writes with floats on Geforce GTX 480}

\section{Results}	
	\subsection{Pure shear of a cube}
	\subsection{Compression of a cube}
	\subsection{GPU performance (+comparison of different implementations)}
	\subsection{Simulation of liver deformation}
	\subsection{A medical application: cataract surgery}
